# Figure 3

```{r}
pkgs <- c("fs", "futile.logger", "configr", "stringr", "ggpubr", "ggthemes", 
          "jhtools", "glue", "ggsci", "patchwork", "tidyverse", "dplyr", "Seurat", 
          "scDblFinder", "paletteer", "readxl", "writexl", "ComplexHeatmap", 
          "circlize", "data.table")  
for (pkg in pkgs){
  suppressPackageStartupMessages(library(pkg, character.only = T))
}

fs::dir_create("./results/fig3")

# themes ----
my_theme1 = theme_classic(base_size = 6) + 
  theme(legend.key.size = unit(3, "mm"), axis.text = element_text(color = "black"))
# colors ----
mmu_mph_cols <- c("MM1" = "#1963b3", "MM2" = "#7ca3b8", "MM3" = "#97cebf", 
                  "MM4" = "#789cf2", "MM5" = "#80c187", "MM6" = "#3f8896", 
                  "MM7" = "#28c694")

```

## (a) DEG analysis

Heatmap of up-regulated genes in PGE2 & IL34 group.

```{r}
## fig3a -----
selected_genes <- read_rds("./data/bulk_rna_select_genes.rds")
fpkm5 <- read_csv("./data/wangchao_mouse_v2.csv") %>% dplyr::filter(gene_name %in% selected_genes) %>% as.data.frame() %>% 
  tibble::column_to_rownames("gene_name") %>% dplyr::select(-1)


vst_dat3 <- fpkm5 %>% t() %>% as.data.frame() %>% 
  mutate(trt = rep(c("CTL", "PGE2", "IL34", "PGE2 & IL34"), each = 2)) %>% 
  aggregate(data = ., . ~ trt, FUN = mean) %>% column_to_rownames("trt") %>% 
  scale() %>% t()
gene_labs2 <- c("Ccr1", "Il6", "Cxcl1", "Nos2", "Egln3", "Saa3", "Inhba", 
                "Vcan", "Nt5e", "F7", "Sdc1", "F10", "Trem1")
labs_idx2 <- which(rownames(vst_dat3) %in% gene_labs2)

top_anot1 <- 
  ComplexHeatmap::rowAnnotation(foo = anno_mark(at = labs_idx2, 
                                                labels = rownames(vst_dat3)[labs_idx2], 
                                                labels_gp = gpar(fontsize = 6)))

col_fun = colorRamp2(c(-2, 0, 2), c("#3A488AFF", "white", "#BE3428FF"))
htp1 <- ComplexHeatmap::Heatmap(vst_dat3, show_row_names = F, cluster_columns = F, 
                                # row_names_gp = gpar(fontsize = 4),
                                right_annotation = top_anot1, 
                                column_names_gp = gpar(fontsize = 8,angle = 45), 
                                name = "Z-score", col = col_fun, 
                                height = unit(8, "cm"), width = unit(2, "cm"))
pdf("./results/fig3/fig3a_mks_heatmap.pdf", width = 3, height = 4)
print(htp1)
d <- dev.off()

```

```{r echo=FALSE, out.width = '800px', out.height='600px', fig.align='center'}
print(htp1)
```

## (b) GSEA plot of MM1 driver genes

```{r}
## fig3b: correlated driver genes gsea ----
trm2gn2 <- read_rds("./data/trm2gn2_bulk_common_genes.rds")
my_theme1 = ggplot2::theme_classic(base_size = 8) + 
  theme(legend.key.size = unit(3, "mm"), axis.text = element_text(color = "black", size = 8))

mks_df <- read_csv("./data/all_to_MM1_lineage_drivers_v2.csv")

inpt2 <- structure(mks_df[[2]], names = mks_df[[1]])

set.seed(241101)
tst_res <- clusterProfiler::GSEA(inpt2, TERM2GENE = trm2gn2, minGSSize = 1, 
                                 pvalueCutoff = 1, pAdjustMethod = "fdr", seed = T)
pval <- tst_res@result$pvalue %>% signif(., 4)
nes <- tst_res@result$NES %>% round(digits = 3)
tst_res@result %>% write_csv("bulk_feats_corr_lineage_gsea_res.csv")
gsea_p1 <- 
  enrichplot::gseaplot2(tst_res, geneSetID = 1, subplots = 1:2, pvalue_table = F, 
                        color = "#11aa11FF", base_size = 7, 
                        title = glue("Mouse MM1 Mph: P = {pval} | NES = {nes}"))
gsea_p1[[1]] <- gsea_p1[[1]] +   
  theme(plot.title = element_text(size = 8, hjust = .5))
ggsave("./results/fig3/fig3b_bulk_feats_corr_lineage_gsea_dot.pdf", 
       gsea_p1, width = 7, height = 6, unit = "cm")

```

```{r echo=FALSE, fig.width = 7, fig.height = 6}
gsea_p1
```

### Fig.S2ab Cell fate analysis of mouse macrophage

UMAP of pseudotime and fate probability

```{r}
# 10x 3' Mph data
mmu_mph <- read_rds("./data/mph_mmu_seu_anot_fin.rds")

## figS2ab: pseudotime and fate probability plot -----
feat1 <- Seurat::FeaturePlot(mmu_mph, features = "monocle3_pseudotime", pt.size = 1e-2) + 
  my_theme1 + theme(legend.position = "right", plot.title = element_text(hjust = .5)) + 
  labs(color = "", title = "Monocle3 pseudotime") + 
  paletteer::scale_color_paletteer_c("grDevices::Sunset") + 
  ggplot2::coord_fixed() + Seurat::NoAxes()

feat2 <- Seurat::FeaturePlot(mmu_mph, features = "fate_prob_MM1", pt.size = 1e-2) + 
  my_theme1 + theme(legend.position = "right", plot.title = element_text(hjust = .5)) + 
  labs(color = "", title = "Fate probability to MM1") + 
  paletteer::scale_color_paletteer_c("grDevices::Plasma") + 
  ggplot2::coord_fixed() + Seurat::NoAxes()
p <- feat1 | feat2
ggsave("./results/fig3/figS2ab_monocle3_cellrank2.pdf", p, width = 6, height = 3)

```

```{r echo=FALSE, fig.width = 6, fig.height = 3}
p
```

### Fig.S2c Heatmap of fate genes

```{r}
## figS2c: specific gene heatmap ----
genes = c("Arg1", "Spp1", "Gapdh", "Basp1", "Vegfa", "Il1a", "Ptgs2", 
          "Hilpda", "Cxcl2", "Slc2a1", "Inhba", "Prdx6", "Egln3", 
          "Mt2", "Clec4d", "Cxcl3", "Ndrg1", "Bnip3", "Ero1l", "Srgn")
pt.matrix <- LayerData(mmu_mph[["RNA"]], layer = "data") %>% 
  .[genes, ] %>% as.matrix() %>% t() %>% as.data.frame() %>% 
  mutate(order = mmu_mph$monocle3_pseudotime, anot_fin = mmu_mph$anot_fin, 
         fate_prob = mmu_mph$fate_prob_MM1) %>% 
  dplyr::filter(!is.infinite(order)) %>%
  dplyr::slice_sample(prop = .1, by = anot_fin) %>%
  dplyr::arrange(fate_prob)  
pt.matrix2 <- pt.matrix[, 1:20] %>% t() 
genes <- rownames(pt.matrix2)
barcodes <- colnames(pt.matrix2)
pt.matrix2 <- t(apply(pt.matrix2, 1, function(x){smooth.spline(x, df=3)$y}))
pt.matrix2 <- t(apply(pt.matrix2, 1, function(x){(x - mean(x))/sd(x)}))

rownames(pt.matrix2) <- genes
colnames(pt.matrix2) <- barcodes
col_func1 <- circlize::colorRamp2(breaks = c(0, 16), colors = c("white", "#cc6611"))
col_func2 <- circlize::colorRamp2(breaks = c(0, 1), colors = c("white", "darkred"))

top_anot1 <- 
  ComplexHeatmap::HeatmapAnnotation(pseudotime = pt.matrix$order[barcodes], 
                                    fate_prob = pt.matrix$fate_prob[barcodes], 
                                    anot = pt.matrix$anot_fin[barcodes], 
                                    col = list(pseudotime = col_func1, 
                                               fate_prob = col_func2, 
                                               anot = mmu_mph_cols))
htkm <- ComplexHeatmap::Heatmap(
  pt.matrix2,
  name = "z-score",
  show_row_names = TRUE,
  cluster_rows = T, cluster_columns = F, 
  show_column_names = FALSE,
  row_names_gp = gpar(fontsize = 7),
  row_title_rot = 0,
  cluster_row_slices = FALSE, 
  top_annotation = top_anot1, raster_by_magick = T, 
  height = nrow(pt.matrix2)*unit(3, "mm"), 
  width = ncol(pt.matrix2)*unit(.05, "mm"))
pdf("./results/fig3/figS2c_gene_fate_prob_heatmap.pdf", width = 4, height = 5)
print(htkm)
d <- dev.off()
```

```{r echo=FALSE}
print(htkm)
```

## (c) Identified macrophage subtypes in bulk RNA data

Dotplot of macrophage subtypes score

```{r}
## fig3c: ----
gsea_dot_func1 <- function(mk_df = NULL, mph_names = NULL, fn = NULL){
  gsea_lst <- lapply(mph_names, \(nm){
    sel_mks <- mk_df %>% dplyr::filter(cluster == nm)
    inpt <- structure(sel_mks$avg_log2FC, names = sel_mks$gene) %>% sort() %>% rev()
    clusterProfiler::GSEA(inpt, TERM2GENE = trm2gn2, minGSSize = 1, 
                          pvalueCutoff = 1, pAdjustMethod = "fdr", seed = T)
  })
  names(gsea_lst) <- mph_names
  gsea_res <- lapply(mph_names, \(nm){
    gsea_lst[[nm]]@result %>% mutate(celltype = nm)
  }) %>% bind_rows() %>% dplyr::arrange((NES)) %>% 
    mutate(celltype = fct(as.character(celltype))) %>% 
    mutate(counts = str_split(core_enrichment, "/") %>% unlist() %>% length())
  write_csv(gsea_res, glue("fig3b_bulk2sce_gsea_res.csv"))
  dot2 <- ggplot2::ggplot(gsea_res, aes(x = NES, y = celltype, size = -log10(p.adjust), color = NES)) + 
    geom_point() + labs(y = "") + my_theme1 + 
    ggplot2::scale_x_continuous(limits = c(-4, 4)) + 
    paletteer::scale_color_paletteer_c("grDevices::Blue-Red 3") + 
    theme(legend.position = "bottom")
  return(dot2)
}

mmu_mph_mks <- read_csv("./data/mph_mmu_anot_fin_mks.csv")
p <- gsea_dot_func1(mk_df = mmu_mph_mks, mph_names = unique(mmu_mph_mks$cluster), fn = "3samp_mks")
ggsave("./results/fig3/fig3c_bulk2sce_gsea_dot.pdf", p, width = 6, height = 6, unit = "cm")

```

```{r echo=FALSE, fig.width = 6, fig.height = 6, fig.align='center'}
p
```

### Fig.S2d Function of MM1

3' Macrophage MM1 markers enrichment dotplot

```{r}
# figS2d
erch_res <- read_csv("./data/mph_mm1_kegg_erch_res.csv")
tst1 <- erch_res %>% dplyr::filter(grepl("Environmental", category)) %>% 
  dplyr::slice_min(order_by = p.adjust, n = 15) %>% tibble %>% 
  mutate(pth_gene_num = str_split(BgRatio, "/", simplify = T)[, 1] %>% as.numeric()) %>% 
  mutate(ratio = Count / pth_gene_num) %>% dplyr::arrange(desc(Count)) %>% 
  dplyr::slice_max(order_by = Count, n = 10) %>% 
  mutate(background = as.numeric(str_split(GeneRatio, "/", simplify = T)[, 2])) %>% 
  mutate(x = Count/background)

dot2 <- tst1 %>% dplyr::arrange(desc(x)) %>% 
  mutate(Description = str_wrap(Description, width = 30) %>% fct() %>% fct_rev(), 
  ) %>% 
  ggplot2::ggplot(aes(x = x, y = Description, 
                      color = -log10(p.adjust), size = Count)) + 
  ggplot2::geom_point() + 
  ggplot2::scale_size(range = c(2, 6), breaks = c(10, 20, 30, 40), labels = c(10, 20, 30, 40)) + 
  my_theme1 + paletteer::scale_color_paletteer_c("ggthemes::Red") + 
  theme(axis.text = element_text(size = 7)) + labs(x = "Gene ratio", y = "") + 
  theme(legend.position = "bottom", axis.line = element_blank(), 
        panel.border = element_rect(fill = NA, color = "black", linewidth = .5), 
        plot.title = element_text(size = 8, hjust = .5)) + 
  labs(title = "Environmental information processing")
ggsave("./results/fig3/figS2d_MM1_kegg_enrich_dot2.pdf", dot2, width = 10, height = 8, unit = "cm")

```

```{r echo=FALSE, fig.width = 10, fig.height = 8, fig.align='center'}
dot2
```
