# Figure 1

Package load and plot settings.

```{r warning=FALSE}
pkgs <- c("fs", "configr", "stringr", "ggpubr", "ggthemes", 
          "jhtools", "glue", "ggsci", "patchwork", "tidyverse", "dplyr", "Seurat", 
          "scDblFinder", "paletteer", "readxl", "writexl", "ComplexHeatmap", 
          "circlize", "data.table", "survminer", "survival")  
for (pkg in pkgs){
  suppressPackageStartupMessages(library(pkg, character.only = T))
}

fs::dir_create("./results/fig1_s1")

# themes -----
my_theme1 <- theme_classic(base_size = 8) + 
  theme(legend.key.size = unit(3, "mm"), axis.text = element_text(color = "black"))
my_theme2 <- theme_classic(base_size = 8) + 
  theme(legend.key.size = unit(3, "mm"), legend.position = "none",
        axis.text = element_text(color = "black"),
        axis.text.x = element_text(angle = 10, vjust = 1, hjust = .5),
        axis.line = element_blank(), 
        axis.ticks = element_line(color = "black"),
        panel.border = 
          element_rect(fill = NA,colour = "black", linewidth = .5, 
                       linetype = "solid"))
my_theme3 <- theme_classic(base_size = 8) + 
  theme(panel.grid.major = element_line(colour = NA),
        panel.background = element_blank(),
        plot.background = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        strip.clip = "off",
        legend.position = 'none',  
        axis.text.y = element_blank(),
        axis.text.x = element_blank(),
        axis.title.y = element_text(size = 8, color = "black"),
        axis.line.x = element_blank(),
        axis.line.y = element_blank(),
        strip.background = element_blank(),
        strip.text.y.left = element_blank(),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank()
  )
color_cells <- c("Acinar cells" = "#BE96BE", "B cells" = "#e8d056", "CAFs" = "#20B2AA", "DC" = "#FFA500",
                 "Ductal cells" = "#9370DB", "Endocrine cells" = "#98FB98", "Endothelial cells" = "#F08080", "Macrophages" = "#1E90FF",
                 "Mast cells" = "#7CFC00", "Neutrophils" = "#A6CEE3", "Stellate cells" = "#808000", "T/NK" = "#FF00FF")
tnk_cols <- c("NK" = "#EE6A50", "CD4T" = "#CB8A93", "CD8T" = "#F2AD88")
grp_cols <- c("Adjacent tissue" = "#2f4a99ff", "Tumor" = "#b92923ff")

mmu_cols <- c("B cells" = "#e8d056", "CAFs" = "#20B2AA", "DCs" = "#FFA500",
              "Ductal cells" = "#9370DB", "Macrophage" = "#1E90FF",
              "Mast cells" = "#7CFC00", "Neutrophil" = "#A6CEE3", "T/NKs" = "#FF00FF", 
              "NK" = "#EE6A50", "CD4T" = "#CB8A93", "CD8T" = "#F2AD88", "gdT" = "#D854AA")


```

## (a) UMAP of all cell types

```{r cache=FALSE}
hsa_all <- read_rds("./data/seu_all_fil1.rds")
Idents(hsa_all) <- hsa_all$anot_modified1 %>% as.character() %>% 
  fct(., levels = c("Acinar cells", "Ductal cells",
                    "CAFs", "Stellate cells",
                    "Endothelial cells", "Endocrine cells",
                    "Mast cells", "Macrophages", "Neutrophils",
                    "DC", "B cells", "T/NK"))

p1 <- DimPlot(hsa_all, reduction = "rpca_umap", group.by = "anot_modified1",
              cols = color_cells, raster = T, pt.size = 1.5) + my_theme1 + 
  theme(plot.title = element_blank(), 
        legend.text = element_text(size = 7), legend.position = "right") + 
  ggplot2::coord_fixed() + Seurat::NoAxes() 
ggsave("./results/fig1_s1/fig1a_rpca_umap_all_ctypes.pdf", 
       p1, width = 7, height = 4.8, unit = "cm")

```

```{r echo=FALSE, fig.width = 7, fig.height = 4.8}
p1
```

### Fig.S1a Cell type markers

Dotplot of cell type markers

```{r cache=FALSE, warning=FALSE, message=FALSE}
endocrine_mks <- c("CHGB", "GCG", "SST", "DPP6", "KCNMB2")
ductal_mks <- c("KRT19", "MMP7", "KRT7", "KRT8")
acinar_mks <- c("AMBP", "CFTR", "PRSS1", "CTRB1", "CTRB2")
stellate_mks <- c("RGS5", "ACTA2", "ADIRF", "PDGFRB")
caf_mks <- c("LUM", "COL1A1", "COL1A2", "DCN")
endo_mks <- c("CDH5", "PLVAP", "VWF", "CLDN5")
mph_mks <- c("AIF1", "CD68", "CD14", "FCGR1A", "CD163")
dc_mks <- c("FLT3", "CD1C", "XCR1", "CLEC9A", "BATF3")
neutro_mks <- c("CSF3R", "S100A9", "S100A8", "MNDA", "LST1")
mast_mks <- c("KIT", "CPA3", "TPSB2")
b_mks <- c("MS4A1", "CD79A", "CD19", "CD79B")
t_mks <- c("CD2", "CD3D", "CD3E", "NKG7", "CD8A", "CD4")

mks_lst1 <- c(acinar_mks, ductal_mks, caf_mks, stellate_mks, 
              endo_mks, endocrine_mks, mast_mks, mph_mks, neutro_mks, 
              dc_mks, b_mks, t_mks)
dot1 <- Seurat::DotPlot(hsa_all, features = mks_lst1, dot.scale = 2.8) + 
  my_theme1 + Seurat::RotatedAxis() + 
  theme(axis.text = element_text(size = 8), 
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + 
  viridis::scale_color_viridis() + 
  labs(x = "", y = "") & theme(legend.position = "bottom") & 
  scale_size(guide = guide_bins(direction = "horizontal"), range = c(.5, 3)) 
ggsave("results/fig1_s1/figS1a_hsa_all_dot.pdf", dot1, width = 16, height = 8, unit = "cm")

```

```{r echo=FALSE, fig.width = 16, fig.height = 8}
dot1
```

## (b) Cell type proportion

```{r message=FALSE, cache=FALSE}
df <- hsa_all@meta.data[, c("sample_id", "anot_modified1", "type2")] %>% 
  as_tibble(rownames = "barcode")
df_ctype <- df %>% group_by(type2, anot_modified1) %>% dplyr::summarise(nc = n()) %>%
  group_by(type2) %>% dplyr::mutate(nt = sum(nc), ratio = nc/nt) %>% ungroup()

df_ctype$anot_modified1 <- factor(df_ctype$anot_modified1,
                                  levels = c("Acinar cells", "Ductal cells",
                                             "CAFs", "Stellate cells",
                                             "Endothelial cells", "Endocrine cells",
                                             "Mast cells", "Macrophages", "Neutrophils",
                                             "DC", "B cells", "T/NK"))
a <- ggplot(data = df_ctype, mapping = aes(x = type2, fill = anot_modified1, y = nc))+
  geom_bar(stat = "identity", width = 0.5, position = 'fill')+
  scale_fill_manual(values = color_cells)+
  scale_y_continuous(labels = scales::percent)+
  guides(fill=guide_legend(title="cell type"))+
  labs(x = "", y = "")+
  my_theme2 
ggsave("./results/fig1_s1/fig1b_bar.pdf", a, width = 4, height = 5, unit = "cm")

```

```{r echo=FALSE, fig.width = 4, fig.height = 5, fig.align='center'}
a
```

Proportion change between adjacent tissue and tumor

```{r warning=FALSE, message=FALSE, cache=FALSE}
df_ctype$type2 <- factor(df_ctype$type2, levels = c("Adjacent tissue", "Tumor"))
df_sam <- df %>% group_by(sample_id, type2, anot_modified1) %>%
  dplyr::summarise(nc = n()) %>% group_by(sample_id, type2) %>%
  dplyr::mutate(nt = sum(nc), ratio = nc/nt) %>% ungroup()

stat_test <- df_sam %>%
  group_by(anot_modified1) %>% 
  rstatix::wilcox_test(ratio ~ type2, p.adjust.method = "none", paired = F)
stat_test <- stat_test %>% 
  mutate(p.adj.signif = case_when(p >= 0.05 ~ "ns",
                                  p >= 0.01 & p < 0.05 ~ "*",
                                  p >= 0.001 & p < 0.01 ~ "**",
                                  p >= 0.0001 & p < 0.001 ~ "***",
                                  p < 0.0001 ~ "****",
                                  TRUE ~ "ns"))
stat_test <- stat_test %>%
  rstatix::add_xy_position(x = "anot_modified1", dodge = 0.9, fun = "median_iqr")
stat_test$y.position <- 0.3

stat_test <- stat_test %>% 
  dplyr::filter(anot_modified1 %in% c("B cells", "DC", "Macrophages",
                                      "Neutrophils", "Mast cells",
                                      "T/NK"))
df_ctype1 <- df_ctype %>% 
  dplyr::filter(anot_modified1 %in% c("B cells", "DC", "Macrophages",
                                      "Neutrophils", "Mast cells",
                                      "T/NK"))
df_ctype1$anot_modified1 <- as.character(df_ctype1$anot_modified1)
df_ctype1$anot_modified1 <- 
  factor(df_ctype1$anot_modified1, 
         levels = c("B cells", "DC", "Macrophages", "Neutrophils", 
                    "Mast cells", "T/NK"))

p1 <- ggplot(df_ctype1, aes(type2, ratio))+
  geom_line(aes(group = anot_modified1), size = .5) +
  geom_point(aes(color = anot_modified1), size = 1.5)+
  scale_color_manual(values = color_cells)+
  geom_rangeframe(sides = "bl") + 
  scale_y_continuous(expand = expansion(mult = c(0.2, 0.3)))+
  scale_x_discrete(expand = expansion(mult = c(0.1)))+
  my_theme3 + labs(x = '', y = '')+
  facet_wrap(~anot_modified1, ncol=1, strip.position = "left")

p2 <- p1 +
  stat_pvalue_manual(stat_test, tip.length = 0.05, 
                     xmin="group1", xmax="group2",
                     position = position_dodge(0), vjust = 0.5, remove.bracket = F,
                     label = "p.adj.signif", size = 3, hide.ns =T)
ggsave("./results/fig1_s1/fig1b_point_line.pdf", p2, width = 2, height = 6, unit = "cm")

```

```{r echo=FALSE, fig.width = 2, fig.height = 6, fig.align='center'}
p2
```

## (c) T/NK cells

Boxplot of T/NK cell frequency in tumor and adjacent tissue

```{r warning=FALSE, message=FALSE}
meta_hsa_all <- read_rds("./data/metadata_new.rds")
cd45_all <- meta_hsa_all %>% dplyr::select(all_of(c("anot_modified1", "type2", "sample_id"))) %>% 
  dplyr::filter(anot_modified1 %in% c("Macrophages", "B cells", "T/NK", "Neutrophils", 
                                      "DC", "Mast cells")) %>% 
  group_by(sample_id) %>% dplyr::add_count() %>% dplyr::rename("sum_n" = "n") %>% 
  dplyr::ungroup() %>% dplyr::distinct() %>% dplyr::rename("cd45_sum" = "sum_n") %>% .[, 3:4]

tnk_new <- read_rds("./data/tnk_anot_fin.rds")
df2 <- tnk_new@meta.data %>% rownames_to_column("barcode") %>% 
  mutate(main_type = "T/NK") %>% 
  dplyr::group_by(sample_id, main_type) %>% dplyr::add_count() %>% 
  dplyr::rename('sum_n' = "n") %>% 
  dplyr::group_by(sample_id, type2, sum_n, anot_fin) %>% 
  dplyr::count()
df3 <- cd45_all %>% left_join(x = ., y = df2) %>% 
  dplyr::rename("t_cell_sum" = "sum_n") %>% mutate(cd45_ratio = n / cd45_sum) %>% 
  dplyr::filter(!is.na(type2))
box1 <- ggplot2::ggplot(df3, aes(x = type2, y = cd45_ratio, fill = type2)) + 
  ggplot2::geom_boxplot( outlier.size = 0, 
                         width = .5, outlier.alpha = 0) + 
  ggplot2::stat_summary(geom = "line", linewidth = .5, fun = "mean", color = 'black', group = "type2") + 
  ggplot2::stat_summary(geom = "point", shape = 21, fill = "#cFcF00FF", fun = "mean") + 
  ggplot2::facet_wrap(~ anot_fin, nrow = 1) + 
  # theme_classic(base_size = 6) + 
  scale_fill_manual(values = grp_cols) + 
  my_theme2 + theme(axis.title = element_text(size = 8), 
                    axis.ticks.x = element_blank(), 
                    axis.text.x = element_blank()) + 
  labs(x = "", y = "Ratio of CD45+ cells") + 
  ggpubr::stat_compare_means(aes(group = type2), method = "wilcox.test",
                             hide.ns = F, size = 2, label = "p.signif",
                             bracket.size = .3, label.y = .8, 
                             comparisons = list(c("Adjacent tissue", "Tumor"))) +
  ggplot2::scale_y_continuous(limits = c(0, .9), breaks = c(0.0, 0.2, 0.4, .6, 0.8))
ggsave("./results/fig1_s1/fig1c_tcell_cd45_ratio_boxplot.pdf", box1, width = 6, height = 4, unit = "cm")

```

```{r echo=FALSE, fig.width = 6, fig.height = 4}
box1
```

### Fig.S1b UMAP of T/NK cells

```{r warning=FALSE}
# figS1b: 4*4 -----
dim1 <- Seurat::DimPlot(tnk_new, reduction = "cca4umap",  
                        raster = T, pt.size = 1.5) + 
  my_theme1 + ggplot2::coord_fixed() + Seurat::NoAxes() + 
  theme(legend.position = "right", text = element_text(size = 8)) + 
  labs(title = "") + 
  scale_color_manual(values = tnk_cols)
ggsave("./results/fig1_s1/figS1b_tnk_umap_dim1.pdf", dim1, width = 7, height = 6, unit = "cm")

```

```{r echo=FALSE, fig.width = 7, fig.height = 6}
dim1
```

### Fig.S1c T/NK cell markers

Dotplot of T/NK cell markers

```{r warning=FALSE}
# figS1c: 3*4 -----
tnk_feats <- c("CD2", "CD3D", "CD3E", "CD4", "FOXP3", "CCR7", 
               "CD8A", "CD8B", "GZMK", "NKG7", "KLRD1", "GZMB")
dot2 <- Seurat::DotPlot(tnk_new, features = tnk_feats, dot.scale = .5) + 
  my_theme1 + Seurat::RotatedAxis() + 
  theme(axis.text = element_text(color = "black")) + 
  viridis::scale_color_viridis() + labs(x = "", y = "") + 
  theme(legend.position = "right") + 
  scale_size(range = c(.5, 3))
ggsave("./results/fig1_s1/figS1c_tnk_dotplot.pdf", dot2, width = 9, height = 5, unit = "cm")

```

```{r echo=FALSE, fig.width = 9, fig.height = 5}
dot2
```

## (d) Cytotoxicity score of T/NK cells

```{r warning=FALSE}
ssgsea_score <- read_rds("./data/tcellsi_ssgsea_score.rds")
ssgsea_score <- ssgsea_score[, colnames(tnk_new)]
df4 <- ssgsea_score %>% t() %>% as.data.frame() %>% 
  mutate(type2 = tnk_new$type2, anot = tnk_new$anot_modified_241020, 
         sample_id = tnk_new$sample_id) 
box2 <- ggplot2::ggplot(df4, aes(x = type2, y = Cytotoxicity, fill = type2)) + 
  ggplot2::geom_boxplot( outlier.size = 0, 
                         width = .5, outlier.alpha = 0) + 
  ggplot2::stat_summary(geom = "line", linewidth = .5, fun = "mean", color = 'black', group = "type2") + 
  ggplot2::stat_summary(geom = "point", shape = 21, fill = "#cFcF00FF", fun = "mean") + 
  ggplot2::facet_wrap(~ anot, nrow = 1) + 
  theme_classic(base_size = 6) + 
  scale_fill_manual(values = grp_cols) + 
  my_theme2 + theme(axis.ticks.x = element_blank(), 
                    axis.text.x = element_blank(), 
                    legend.position = 'bottom') + 
  labs(x = "", y = "Cytotoxicity score") + 
  ggpubr::stat_compare_means(aes(group = type2), method = "wilcox.test",
                             hide.ns = F, size = 2, label = "p.signif",
                             label.y.npc = 1.2,
                             bracket.size = .3, 
                             comparisons = list(c("Adjacent tissue", "Tumor"))) +
  ggplot2::scale_y_continuous(limits = c(0.2, 1.3),
                              breaks = c(0.2, 0.4, 0.6, 0.8, 1, 1.2))
ggsave("./results/fig1_s1/fig1d_tcell_type_cytotoxic_boxplot.pdf", box2, width = 6, height = 5, unit = "cm")

```

```{r echo=FALSE, fig.width = 6, fig.height = 5}
box2
```

## (e) Macrophage subtype

UMAP of macrophage subtypes

```{r warning=FALSE}
## colors ----
hsa_mph_cols <- c("HM1" = "#1963b3", "HM2" = "#7ca3b8", "HM3" = "#97cebf", 
                  "HM4" = "#789cf2", "HM5" = "#8071e7", "HM6" = "#bf88d6")

# fig1e: 4*4 cm^2 -----
hsa_mph <- read_rds("./data/mph_hsa_seu_anot_fin.rds")

dim1 <- Seurat::DimPlot(hsa_mph, reduction = "cca2umap", pt.size = 1.5, raster = T, 
                        group.by = "anot_fin") + my_theme1 + 
  ggplot2::coord_fixed() + scale_color_manual(values = hsa_mph_cols) + 
  Seurat::NoAxes() + labs(title = "") + theme(legend.position = "right")
ggsave("./results/fig1_s1/fig1e_hsa_mph_dim1.pdf", dim1, width = 5, height = 4.8, unit = "cm")

```

```{r echo=FALSE, fig.width = 5, fig.height = 4.8}
dim1
```

### Fig.S1d Macrophage subtype markers

Heatmap of macrophage subtype markers

```{r warning=FALSE, message=FALSE}
# figS1d: -----
mks_hm1 <- c("HILPDA", "SLC2A1", "EGLN3", "HK2", "NDRG1", "INHBA")
mks_hm2 <- c("SELENOP", "IGF1", "C1QC", "C1QB", "C1QA", "HGF")
mks_hm3 <- c("GTSE1", "CDC20", "MKI67", "TOP2A", "UBE2C", "RRM2")
mks_hm4 <- c("EGFR", "ITGA2", "ARHGEF28", "PARD3B", "PCDH7")
mks_hm5 <- c("FCN1", "IFIT1", "IFIT2", "S100A8", "IL1B", "PTGS2")
mks_hm6 <- c("CCL18", "ACP5", "APOC1", "FABP5", "APOE")
mks_hm_lst <- list(hm1 = mks_hm1, hm2 = mks_hm2, hm3 = mks_hm3, hm4 = mks_hm4, 
                   hm5 = mks_hm5, hm6 = mks_hm6)
names(mks_hm_lst) <- str_to_upper(names(mks_hm_lst))

avg_exps <- 
  Seurat::AverageExpression(hsa_mph, assays = "RNA", features = unlist(mks_hm_lst)) %>% 
  .[[1]] %>% as.matrix() %>% t() %>% scale()
df1 <- avg_exps %>% as.data.frame() %>% rownames_to_column("type") %>% 
  pivot_longer(cols = -1, names_to = "gene", values_to = "expr") %>% 
  mutate(gene = fct(as.character(gene), levels = unlist(mks_hm_lst)), 
         type = fct(as.character(type), levels = paste0("HM", 1:6)))
tile1 <- ggplot2::ggplot(df1, aes(x = gene, y = type, fill = expr)) + 
  ggplot2::geom_tile() + my_theme1 + Seurat::RotatedAxis() + 
  paletteer::scale_fill_paletteer_c("grDevices::Viridis") + 
  labs(y = "", x = "", fill = "Z-score") + ggplot2::coord_flip() + 
  theme(legend.position = "right", axis.line = element_blank(), 
        axis.ticks.x = element_blank(), axis.text = element_text(size = 8), 
        panel.border = element_rect(fill = NA, color = "black", linewidth = .5))
ggsave("./results/fig1_s1/figS1d_hsa_mph_sel_mks_heatmap1.pdf", tile1, width = 7, height = 10, unit = "cm")
```

```{r echo=FALSE, fig.width = 7, fig.height = 10}
tile1
```

## (f) Macrophage subtype proportion

```{r}
## fig1f: bar and lines ---
cnt_df <- hsa_mph@meta.data %>% dplyr::select(all_of(c("anot_fin", "type2"))) %>% 
  group_by(anot_fin, type2) %>% dplyr::count()
bar2 <- ggplot2::ggplot(cnt_df, aes(x = type2, y = n, fill = anot_fin)) + 
  geom_bar(stat = 'identity', position = "fill", width=0.5) + 
  scale_fill_manual(values = hsa_mph_cols) + 
  scale_y_continuous(labels = scales::percent) +
  guides(fill = guide_legend(title="cell type")) +
  labs(x = "", y = "") + my_theme2
ggsave("./results/fig1_s1/fig1f_mph_hsa_anot_fin_barplot.pdf", bar2, width = 4, height = 5, unit = "cm")

```

```{r echo=FALSE, fig.width = 4, fig.height = 5, fig.align='center'}
bar2
```

Proportion change between adjacent tissue and tumor

```{r}
## lines ---
cnt_df2 <- hsa_mph@meta.data %>% 
  dplyr::select(all_of(c("type2", "anot_fin", "sample_id"))) %>% 
  group_by(sample_id) %>% dplyr::add_count() %>% dplyr::rename('sum_n' = "n") %>% 
  ungroup() %>% dplyr::group_by(sample_id, anot_fin) %>% dplyr::add_count() %>% 
  dplyr::rename("celtp_n" = "n") %>% mutate(prop = celtp_n/sum_n) %>% dplyr::distinct()
p1 <- cnt_df2 %>% #dplyr::filter(anot_fin == "HM5") %>% 
  ggplot(., aes(x = type2, y = prop))+
  stat_summary(fun = "mean", geom = "line", linewidth = .5, 
               group = "anot_fin", color = "black") +
  stat_summary(fun = "mean", geom = "point", size = 1.5,
               mapping = aes(color = anot_fin)) +
  scale_color_manual(values = hsa_mph_cols)+
  geom_rangeframe() +   # 坐标轴分离
  scale_x_discrete(expand = expansion(mult = c(0.1))) +
  my_theme3 +
  labs(x = '', y = '') +
  facet_wrap(~ anot_fin, ncol=1, strip.position = "left") + 
  ggpubr::stat_compare_means(aes(group = type2), 
                             # comparisons = list(c("Adjacent tissue", "Tumor")),
                             method = "wilcox.test", size = 2, 
                             hide.ns = T, label = "p.signif", 
                             label.y = .6, label.x = 1.43)
ggsave("./results/fig1_s1/fig1f_mph_hsa_prop_line.pdf", p1, width = 2, height = 6, unit = "cm")

```

```{r echo=FALSE, fig.width = 2, fig.height = 6, fig.align='center'}
p1
```

## (g) Macrophage subtype scores of TCGA data

HM1 score

```{r message=FALSE, warning=FALSE}
## colors -----
color <- c("High" = "#B92923", "Low" = "#2F4A99")

# fig1g: 6*5 ----
TCGA_tide_m1 <- read_csv(glue("./data/TCGA_tide_m1.csv"))
vln.df <- TCGA_tide_m1[, c("Patient", "m1_score", colnames(TCGA_tide_m1)[17:27])]
rownames(vln.df) <- vln.df$Patient
vln.df <- vln.df[, -1]
vln.df <- pivot_longer(vln.df, cols =!m1_score, 
                       names_to = "cluster", values_to = "level")
my_comparisons <- list(c("High", "Low"))
m1 <- min(TCGA_tide_m1[, "m1_score"])
m2 <- max(TCGA_tide_m1[, "m1_score"])

vln.df <- vln.df %>% 
  dplyr::filter(., cluster %in% c("CD8_level", "Dysfunction_level", 
                                  "Exclusion_level")) %>% 
  mutate(cluster = case_when(cluster == "Dysfunction_level" ~ "Dysfunction", 
                             cluster == "Exclusion_level" ~ "Exclusion", 
                             TRUE ~ cluster)) %>% 
  mutate(cluster = gsub("_", " ", cluster)) %>% 
  mutate(level = factor(level, level = c("Low", "High")))

vln4 <- vln.df %>% ggplot(aes(level,m1_score))+
  ggplot2::geom_violin(aes(fill = level), scale = "width", ) +
  geom_boxplot(width = 0.2, outlier.size = 0, outlier.stroke = NA)+
  facet_wrap(~cluster) +
  scale_fill_manual(values=color) +
  scale_x_discrete("") +
  scale_y_continuous(limits =c(m1,m2+0.25*(m2-m1)))+
  my_theme2 + theme(axis.text.x = element_text(angle = 0)) + 
  ylab("HM1 score") +
  ggpubr::stat_compare_means(method = "wilcox.test",size = 2,
                             label.y = m2 + 0.05*(m2-m1),
                             label = "p.signif", 
                             hide.ns = T, tip.length = .05, 
                             comparisons = list(c("Low", "High")))
ggsave("./results/fig1_s1/fig1g_boxplot_HM1_gsva_241101.pdf", vln4, width = 7, height = 3, units = "cm")

```

```{r echo=FALSE, fig.width = 7, fig.height = 3}
vln4
```

### Fig.S1e HM5/HM6 score

```{r warning=FALSE,message=FALSE}
score_vln_func1 <- function(score_df = NULL, col_nm = NULL, label = NULL) {
  color <- c("High" = "#B92923", "Low" = "#2F4A99")
  colnames(score_df) <- 
    gsub(pattern = col_nm, replacement = "hm_score", x = colnames(score_df))
  vln.df <- score_df[,c("Patient", "hm_score", colnames(score_df)[17:27])]
  rownames(vln.df) <- vln.df$Patient
  vln.df <- vln.df[, -1]
  vln.df <- pivot_longer(vln.df, cols = !hm_score, names_to = "cluster", 
                         values_to = "level")
  my_comparisons <- list(c("High", "Low"))
  m1 <- min(score_df[, "hm_score"])
  m2 <- max(score_df[, "hm_score"])
  
  vln.df <- vln.df %>% 
    dplyr::filter(., cluster %in% c("CD8_level", "Dysfunction_level", 
                                    "Exclusion_level")) %>% 
    mutate(level = factor(level, level = c("Low", "High")), 
           cluster = case_when(cluster == "CD8_level" ~ "CD8 level", 
                               cluster == "Dysfunction_level" ~ "Dysfunction", 
                               cluster == "Exclusion_level" ~ "Exclusion"))
  
  vln4 <- vln.df %>% ggplot(aes(level, hm_score))+
    ggplot2::geom_violin(aes(fill = level), scale = "width")+
    geom_boxplot(width = 0.2, outlier.size = 0, outlier.stroke = NA)+
    facet_wrap(~cluster) +
    scale_fill_manual(values=color) +
    scale_x_discrete("") + 
    scale_y_continuous(limits =c(m1, m2 + 0.25*(m2-m1)))+
    my_theme2 + 
    theme(axis.text.x = element_blank(), legend.position = 'right', 
          axis.ticks.x = element_blank()) + 
    ylab(glue("{label} score")) +
    ggpubr::stat_compare_means(method = "wilcox.test",size = 2,
                               label.y=m2+0.05*(m2-m1),
                               label = "p.signif", 
                               hide.ns = T, tip.length = .05, 
                               comparisons = list(c("Low", "High")))
  return(vln4)
}

score_df1 <- read_csv("./data/TCGA_tide_h4.csv")
vln5 <- score_vln_func1(score_df = score_df1, col_nm = "h4_score", label = "HM5")
ggsave("./results/fig1_s1/figS1e_boxplot_HM5_gsva.pdf", vln5, width = 10, height = 3, units = "cm")

```

```{r echo=FALSE, fig.width = 10, fig.height = 3}
vln5
```

HM6 score

```{r warning=FALSE, message=FALSE}
score_df2 <- read_csv("./data/TCGA_tide_h6.csv")
vln6 <- score_vln_func1(score_df = score_df2, col_nm = "h6_score", label = "HM6")
ggsave("./results/fig1_s1/figS1e_boxplot_HM6_gsva.pdf", vln6, width = 10, height = 3, units = "cm")
```

```{r echo=FALSE, fig.width = 10, fig.height = 3}
vln6
```

## (h) Survival analysis of TCGA data

Kaplan-Meier curve stratified by HM1 score

```{r warning=FALSE, message=FALSE}
TCGA_tide_m1_surv <- read_csv("./data/TCGA_tide_m1_surv.csv")
TCGA_tide_m1_surv <- mutate(TCGA_tide_m1_surv, OS.time = OS.time/30)

grid.draw.ggsurvplot <- function(x) { 
  survminer:::print.ggsurvplot(x, newpage = FALSE)
}
fit = survival::survfit(Surv(as.numeric(TCGA_tide_m1_surv$OS.time), 
                             as.numeric(TCGA_tide_m1_surv$OS)) ~ unlist(TCGA_tide_m1_surv[,"m1_score_level"]), 
              data =TCGA_tide_m1_surv)
p1 = survminer::ggsurvplot(fit, data = TCGA_tide_m1_surv,
                           xlab = "Time in Months",
                           legend = c(0.8, 0.8),
                           legend.title = "HM1 score level",
                           legend.labs = c("High","Low"),
                           pval = TRUE, pval.size = 2, 
                           pval.coord = c(60 * 0.6, 0.9),
                           conf.int = F, risk.table = F,
                           font.tickslab = c(7, "plain", "black"),
                           linetype = c("solid", "solid"), # Change line type by groups
                           palette = c("High"="#B92923", "Low"="#2F4A99"),
                           fontsize = 5,
                           xlim = c(0, 60),
                           font.legend = 6,
                           break.time.by = 12,
                           size = .5, 
                           linewidth = .5, 
                           font.x = c(8, "black"),
                           font.y = c(8, "black")) 
ggsave("./results/fig1_s1/fig1h_TCGA_HM1_score_level_KM_OS.pdf", 
       p1, width = 5, height = 4, unit = "cm")
```

```{r echo=FALSE, fig.width = 5, fig.height = 4}
p1
```

### Fig.S1f

```{r warning=FALSE, message=FALSE, fig.width = 5, fig.height = 4}
# figS1f -----
surv_df0 <- read_csv("./data/TCGA_tide_m1_surv.csv")
surv_df1 <- read_csv("./data/TCGA_tide_h4_surv.csv")
surv_df2 <- read_csv("./data/TCGA_tide_h6_surv.csv")

tbl1 <- tibble(surv_df_lst = list(surv_df0, surv_df1, surv_df2), 
               nm_lst = c("HM1", "HM5", "HM6"), 
               numeric_column_lst = c("m1_score_level", "h4_score_level", "h6_score_level"), 
               fig_num_lst = c("tst", "figS1f", "figS1f"))
for(idx in 2:nrow(tbl1)) {
  surv_df <- tbl1[["surv_df_lst"]][idx][[1]]
  nm <- tbl1[["nm_lst"]][idx]
  numeric_column <- tbl1[["numeric_column_lst"]][idx]
  fig_num <- tbl1[["fig_num_lst"]][idx]
  
  surv_df$OS.time <- surv_df$OS.time/30
  fit = survfit(Surv(as.numeric(surv_df[["OS.time"]]), 
                     as.numeric(surv_df[["OS"]])) ~ surv_df[[numeric_column]], 
                data =surv_df)
  p1 = survminer::ggsurvplot(fit, data = surv_df, 
                             xlab = "Time in Months",
                             legend = c(0.8, 0.8),
                             legend.title = glue("{nm} score level"),
                             legend.labs = c("High","Low"),
                             pval = TRUE, pval.size = 2, 
                             pval.coord = c(60 * 0.6, 0.9),
                             conf.int = F, risk.table = F,
                             font.tickslab = c(7, "plain", "black"),
                             linetype = c("solid", "solid"), # Change line type by groups
                             palette = c("High" = "#B92923", "Low"="#2F4A99"),
                             fontsize = 5,
                             xlim = c(0, 60),
                             font.legend = 6,
                             break.time.by = 12,
                             size = .5, 
                             linewidth = .5, 
                             font.x = c(8, "black"),
                             font.y = c(8, "black")) 
  ggsave(glue::glue("./results/fig1_s1/{fig_num}_TCGA_{nm}_score_level_KM_OS_241101a.pdf"), 
         p1, width = 5, height = 5, device = "pdf", unit = "cm")
  print(p1)
}

```

## (g) Mouse single cell data

UMAP of all cell types

```{r}
mmu_mph_cols <- c("MM1" = "#1963b3", "MM2" = "#7ca3b8", "MM3" = "#97cebf", 
                  "MM4" = "#789cf2", "MM5" = "#80c187", "MM6" = "#3f8896", 
                  "MM7" = "#28c694")

# fig1i: 4*4 ----
mmu_all <- read_rds("./data/seu_WT_remove_new.rds")

mmu_dim1 <- Seurat::DimPlot(mmu_all, pt.size = 1.5, raster = T, reduction = "umap_rpca",
                            group.by = "anot_manual1") &
  my_theme1 & Seurat::NoAxes() & theme(legend.position = "right") &
  labs(title = "") & scale_color_manual(values = mmu_cols)
ggsave("./results/fig1_s1/fig1i_mmu_all_cells_dim1_241101a.pdf", mmu_dim1, width = 6, height = 4, unit = "cm")

```

```{r echo=FALSE, fig.width = 6, fig.height = 4}
mmu_dim1
```

### Fig.S1g Mouse cell type markers

Dotplot of cell type markers

```{r warning=FALSE, message=FALSE}
# figS1g: 15.3*8 or 8*12 -----
b_mks <- c("Cd79a", "Cd19", "Cd79b", "Ms4a1")
t_mks <- c("Cd3d", "Cd3e", "Nkg7", "Klrd1", "Cd4", "Cd8a")
mono_mks <- c("Cd14", "Cd44", "Cd16")
mph_mks <- c("Cd68", "Adgre1", "Fcgr1", "Ly6c2", "C1qa", "C1qb")
dc_mks <- c("H2-Oa", "H2-DMb2", "Flt3", "Grk3", "Htr7")
mast_mks <- c("Kit", "Tpsb2", "Cpa3", "Hs6st2", "Mcpt4")
stroma_mks <- c("Col1a1", "Col1a2", "Dcn", "Col5a2", "Bgn")
neutro_mks <- c("Csf3r", "S100a9", "S100a8", "Acod1", "Hdc")
ductal_mks <- c("Krt18", "Krt8",  "Epcam", "Cdkn2a", "Clu")

Idents(mmu_all) <- mmu_all$anot_manual1 %>% as.character() %>% 
  fct(., levels = c("Ductal cells", "CAFs", "DCs", "Mast cells", 
                    "Macrophage", "Neutrophil", "B cells", "T/NKs"))
mks_lst <- fct(c(ductal_mks, stroma_mks, dc_mks, mast_mks, 
                 mph_mks, neutro_mks, b_mks, t_mks))

dot1 <- mmu_all %>% 
  Seurat::DotPlot(., features = mks_lst, dot.scale = 3, assay = "RNA") + 
  my_theme1 + theme(axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)) + 
  viridis::scale_color_viridis() + labs(x = "", y = "") + 
  ggplot2::coord_flip() +
  theme(axis.text = element_text(size = 8), legend.position = "bottom") + 
  ggplot2::scale_size(range = c(.5, 2.5))
ggsave("./results/fig1_s1/figS1g_mouse_all_cells_dotplot2.pdf", dot1, width = 8, height = 12, unit = "cm")

```

```{r echo=FALSE, fig.width = 8, fig.height = 12}
dot1
```

## (j) Mouse macrophage subtype

```{r}
## fig1k: 6*4 inch
rds_fn10 <- 
  "/cluster/home/danyang_jh/projects/panlab/analysis/wangchao/mouse/tenx/merged_new/macrophage/mph_fin/mph_mmu_seu_anot_fin.rds"
mmu_mph <- read_rds("./data/mph_mmu_seu_anot_fin.rds")

dim1 <- Seurat::DimPlot(mmu_mph, pt.size = 1e-3, reduction = "umap", 
                        group.by = "anot_fin", raster = F) + 
  my_theme1 + Seurat::NoAxes() + labs(title = "") + 
  ggplot2::coord_fixed() + 
  theme(legend.position = "right") + scale_color_manual(values = mmu_mph_cols)
ggsave("./results/fig1_s1/fig1j_mmu_mph_dim1.pdf", dim1, width = 6, height = 4, unit = "in")

```

## (k) Human--mouse comparison of TAM clusters

Calculate human TAM markers score on mouse

```{r warning=FALSE, message=FALSE}
# fig1k -----
gsea_res <- read_csv("./data/gsea_res_mph_hsa2mmu.csv")
dot1 <- gsea_res %>% dplyr::arrange(NES) %>% 
  mutate(target = fct(target, levels = paste0("MM", 1:7)), 
         ID = fct(ID, levels = paste0("HM", 1:6))) %>% 
  ggplot2::ggplot(aes(x = ID, y = target, color = NES, size = -log10(pvalue))) + 
  geom_point() + 
  my_theme1 + theme(text = element_text(size = 7)) + 
  ggplot2::scale_colour_gradient2(low = "navy", high = "#f90000", mid = "white") + 
  Seurat::RotatedAxis() + 
  theme(axis.line = element_blank(), axis.text = element_text(color = "black"), 
        panel.border = element_rect(fill = NA, color = "black", linewidth = .5)) + 
  labs(x = "", y = "")
ggsave("./results/fig1_s1/fig1k_human_mouse_mph_gsea_dot1.pdf", dot1, width = 7, height = 6, unit = "cm")

```

```{r echo=FALSE, fig.width = 7, fig.height = 6}
dot1
```

## (l) Markers expression of human-mouse TAM

Human-mouse common markers

```{r warning=FALSE, message=FALSE}
# fig1l: 4*4 inch in total
human <- read_rds("./data/mph_hsa_seu_anot_fin.rds")
mouse <- read_rds("./data/mph_mmu_seu_anot_fin.rds")

Idents(human) <- human$anot_fin
Idents(mouse) <- mouse$anot_fin

sel_gns_hsa <- c("SDC","HILPDA","SLC2A1","EGLN3","HK2","NDRG1","INHBA","MIF","SERPINE1","CXCL3","CXCL2","CXCL1","SAA1","NINJ1","VEGFA",
              "SELENOP","IGF1","C1QC","C1QB","C1QA","GAS6","HGF","AXL",
              "GTSE1","CDC20","DLGAP5","CDCA3","MKI67","TOP2A","UBE2C","RRM2",
              "HMGA2","EGFR","PBX1","ITGA2","SAMD12","ZEB1","MLLT3")
sel_gns_mmu <- c("Sdc","Hilpda","Slc2a1","Egln3","Hk2","Ndrg1","Inhba","Mif","Serpine1","Cxcl3","Cxcl2","Cxcl1","Saa3","Ninj1","Vegfa",
                  "Selenop","Igf1","C1qc","C1qb","C1qa","Gas6","Hgf","Axl",
                  "Gtse1","Cdc20","Dlgap5","Cdca3","Mki67","Top2a","Ube2c","Rrm2",
                  "Hmga2","Egfr","Pbx1","Itga2","Samd12","Zeb1","Mllt3")

sdc_gene_hsa <- c("SDC1","SDC2","SDC3","SDC4")
sdc_gene_mmu <- c("Sdc1","Sdc2","Sdc3","Sdc4")

sdc_expr_hsa <- AverageExpression(human, features = sdc_gene_hsa, assays = "RNA",
                              layer = "data") %>% .[[1]] %>% colMeans()
hsa_mtx <- AverageExpression(human, features = sel_gns_hsa, assays = "RNA",
                             layer = "data") %>% .[[1]] %>% as.matrix() %>% 
  t() %>% as.data.frame() %>% mutate(SDC = sdc_expr_hsa) %>% t() %>% 
  .[sel_gns_hsa[1:31], 1:3] %>% t() %>% scale() %>% t() %>% as.matrix()

sdc_expr_mmu <- AverageExpression(mouse, features = sdc_gene_mmu, assays = "RNA",
                              layer = "data") %>% .[[1]] %>% colMeans()
mmu_mtx <-AverageExpression(mouse, features = sel_gns_mmu, assays = "RNA", 
                            layer = "data") %>% .[[1]] %>% as.matrix() %>% 
  t() %>% as.data.frame() %>% mutate(Sdc = sdc_expr_mmu) %>% t() %>% 
  .[sel_gns_mmu[1:31], 1:3] %>% t() %>% scale() %>% t() %>% as.matrix()
rownames(mmu_mtx) <- gsub("Sdc", "Sdc family", rownames(mmu_mtx))
rownames(hsa_mtx) <- gsub("SDC", "SDC family", rownames(mmu_mtx))

anot_top1 <- ComplexHeatmap::HeatmapAnnotation(species = rep(c("human"), each = 3),
                                               show_annotation_name = F, 
                                               annotation_width = unit(6, "mm"), 
                                               col = list(species = c("human" = "#dd8811", "mouse" = "#1188dd")))
anot_top2 <- ComplexHeatmap::HeatmapAnnotation(species = rep(c("mouse"), each = 3), 
                                               show_annotation_name = F, 
                                               annotation_width = unit(6, "mm"), 
                                               col = list(species = c("human" = "#dd8811", "mouse" = "#1188dd")))

col_func1 <- circlize::colorRamp2(breaks = c(-1, 0, 1), colors = c("blue", 'white', "red"))
htp_hsa <- ComplexHeatmap::Heatmap(hsa_mtx, name = "Z-score", cluster_rows = F, 
                                   cluster_columns = F, show_row_dend = F, 
                                   show_row_names = T, row_names_side = "left", 
                                   row_names_gp = gpar(fontsize = 6), col = col_func1, 
                                   column_names_gp = gpar(fontsize = 7), 
                                   show_heatmap_legend = F, top_annotation = anot_top1)
htp_mmu <- ComplexHeatmap::Heatmap(mmu_mtx, name = "Z-score", cluster_rows = F, 
                                   cluster_columns = F, show_row_dend = F, 
                                   show_row_names = T, row_names_side = "right", 
                                   row_names_gp = gpar(fontsize = 6),  col = col_func1, 
                                   column_names_gp = gpar(fontsize = 7), 
                                   top_annotation = anot_top2)
htp_lst <- htp_hsa + htp_mmu
pdf("./results/fig1_s1/fig1l_hsa_mmu_common_mks_heatmap.pdf", width = 4, height = 4)
ComplexHeatmap::draw(htp_lst)
d <- dev.off()
```

```{r echo=FALSE, message=FALSE}
ComplexHeatmap::draw(htp_lst)
```

### Fig.S1h Mouse TAM markers

Mouse TAM markers expression on 10x 3' platform and 10x 5' platform

```{r warning=FALSE, message=FALSE}
# figS1h: mouse markers of 3' and 5' mouse macrophage ----
mks_mm1 <- c("Inhba", "Egln3", "Hilpda", "Ndrg1", "Slc2a1", "Hk2")
mks_mm2 <- c("Igf1", "Selenop", "C1qc", "C1qa", "C1qb", "Hgf")
mks_mm3 <- c("Rrm2", "Ube2c", "Top2a", "Cdc20", "Mki67", "Gtse1")
mks_mm4 <- c("Itga2", "Egfr", "Arhgef28", "Pard3b", "Pcdh7")
mks_mm5 <- c("Il1r2", "Fgf13", "Ccl17", "Klrb1b", "Ramp3")
mks_mm6 <- c("Ly6c2", "Sell", "Plac8", "Chil3", "Vcan")
mks_mm7 <- c("Ccl5", "Cxcl9", "Gbp2", "Gbp4", "Cxcl10", "Gbp5")
mks_mm_lst <- list(mm1 = mks_mm1, mm2 = mks_mm2, mm3 = mks_mm3, mm4 = mks_mm4,
                   mm5 = mks_mm5, mm6 = mks_mm6, mm7 = mks_mm7)
names(mks_mm_lst) <- str_to_upper(names(mks_mm_lst))

mmu_mph <- read_rds("./data/mph_mmu_seu_anot_fin.rds")
avg_exps <- 
  Seurat::AverageExpression(mmu_mph, assays = "RNA", features = unlist(mks_mm_lst)) %>% 
  .[[1]] %>% as.matrix() %>% t() %>% scale() %>% t() %>% .[unlist(mks_mm_lst), ]

mmu_mph_2samp <- read_rds("./data/seu_2samp.rds")
avg_exps2 <- 
  Seurat::AverageExpression(mmu_mph_2samp, assays = "RNA", features = unlist(mks_mm_lst)) %>% 
  .[[1]] %>% as.matrix() %>% t() %>% scale() %>% t() %>% .[unlist(mks_mm_lst), ]

top_anot1 <- ComplexHeatmap::HeatmapAnnotation(type = rep("10x 3' platform", 7), 
                                               show_annotation_name = F, 
                                               col = list(type = c("10x 3' platform" = "#ff8844")))
htp1 <- ComplexHeatmap::Heatmap(avg_exps, show_heatmap_legend = F, 
                                top_annotation = top_anot1, 
                                cluster_rows = F, cluster_columns = F, 
                                row_names_gp = gpar(fontsize = 4), 
                                column_names_gp = gpar(fontsize = 4))
top_anot2 <- ComplexHeatmap::HeatmapAnnotation(type = rep("10x 5' platform", 7), 
                                               show_annotation_name = F, 
                                               col = list(type = c("10x 5' platform" = "#4488ff")))
htp2 <- ComplexHeatmap::Heatmap(avg_exps2, show_heatmap_legend = T, 
                                top_annotation = top_anot2, name = "Z-score", 
                                cluster_rows = F, cluster_columns = F, 
                                row_names_gp = gpar(fontsize = 4), 
                                column_names_gp = gpar(fontsize = 4))
htp_lst <- htp1 + htp2
pdf("./results/fig1_s1/figS1h_mmu_mph_heatmap.pdf", width = 4, height = 4)
draw(htp_lst)
d <- dev.off() 
```

```{r echo=FALSE, message=FALSE}
draw(htp_lst)
```
