# Preface {.unnumbered}

This is a Quarto book of code accompanying our manuscript "A CD138+ tumor-associated macrophage-Siglec-F+ neutrophil feedforward loop promotes immune evasion in pancreatic cancer".

Immune evasion is a major obstacle in pancreatic cancer therapy. Recent data implicate proinflammatory macrophages in the progression of pancreatic ductal adenocarcinoma (PDAC) and its therapeutic response. However, whether or which of the proinflammatory macrophage subtypes play a crucial role in the immune escape of PDAC remains unclear. Here, we identify a population of CD138+ tumor-associated macrophages (TAMs), characterized by their proinflammatory and neutrophilchemotactic activity, which undergo significant expansion in both patients with PDAC and mouse models. These cells are elicited by a local synergy between IL-34/syndecan-1 and PGE2/EP2 signaling and are associated with immune evasion and poor clinical outcomes in patients, while also promoting immune escape and disease progression in mouse models. Mechanistically, CD138+ TAMs establish a feed-forward loop with immunosuppressive Siglec-F+ neutrophils, which exhibit elevated PGE2 expression, via the secretion of SAA3 and CXCL1. Targeting CD138+ TAMs by disrupting IL-34/syndecan-1 signaling with anti–IL-34 neutralizing antibodies significantly suppressed PDAC progression, especially when combined with anti–PD-1 antibodies. Together, our study elucidates a CD138+ TAM/Siglec-F+ neutrophil axis that drives immune escape in PDAC and proposes a therapeutic strategy that integrates IL-34/syndecan-1 signaling blockade with anti–PD-1 immunotherapy for the treatment of PDAC.

```{r echo=FALSE, eval=FALSE}
pkgs <- c("fs", "configr", "stringr", "ggpubr", "ggthemes", 
          "jhtools", "glue", "ggsci", "patchwork", "tidyverse", "dplyr", "Seurat", 
          "scDblFinder", "paletteer", "readxl", "writexl", "ComplexHeatmap", 
          "circlize", "data.table", "survminer", "survival")  
for (pkg in pkgs){
  suppressPackageStartupMessages(library(pkg, character.only = T))
}
```

```{r}
sessionInfo()
```
